---
title: "Simple Linear Regression"
---

In this session, we will use the same `CASchools` dataset from the previous session and implement a simple linear regression model to explore a question of interest: 

::: {.callout-important appearance="minimal"}
## <i class="codicon codicon-ask" aria-hidden="true" style="font-size:1.5em; vertical-align: middle;"></i> Research Question

<center style="font-size: 1.5em; font-weight: bold;">
Does expenditure per student affect student performance \
in elementary school education?
</center>
:::


## Data Preparation

```{r set-up}
# load packages and dataset
pkgs <- c("tidyverse", "moments", "data.table", "ggsci", "stargazer")
missing <- setdiff(pkgs, rownames(installed.packages()))
if (length(missing) > 0) install.packages(missing)
invisible(lapply(pkgs, function(pkg) suppressPackageStartupMessages(library(pkg, character.only = TRUE))))
f_name <- "https://raw.githubusercontent.com/my1396/FIN5005-Fall2025/refs/heads/main/data/CASchools_test_score.csv"
cas <- read_csv(f_name,
    col_types = cols(
        county = col_factor(), # read as factor
        grades = col_factor()
    )
)
cas %>% as.data.table()
```


Variable of interest: Test scores

- `read` and `math` are average reading and math test scores for each district
- In the [last session](#academic-performance-relationship), we show that these two variables are highly correlated (r=0.92)
- We construct an average test score as the average of the test score for reading and the score of the math test. 

```{r}
# compute TestScore and append it to CASchools
cas <- cas %>%
  mutate(TestScore = (read + math) / 2)
```

## Descriptive Statistics

Refer to the [previous session](#descriptive-analysis-and-data-visualization-with-caschools) for detailed summary statistics and visualizations.

A quick preview 

```{r results="asis"}
v <- setdiff(names(cas), c("district", "school", "county", "grades"))
stargazer(
  as.data.frame(cas[v]), 
  type = ifelse(knitr::is_html_output(), "html", "latex"), 
  digits = 2)
```

## Scatter Plot

Hypothesis: Higher spending per student leads to better test scores.

```{r message=FALSE}
ggplot(cas, aes(x = expenditure, y = TestScore)) +
    geom_point(alpha = 0.6, color = "#1976d2") +
    geom_smooth(method = "lm", color = "#d32f2f", se = FALSE) +
    labs(
        title = "Scatter Plot of Test Scores vs. Expenditure per Student",
        x = "Expenditure per Student (USD)",
        y = "Average Test Score"
    ) +
    theme_minimal(base_size = 14)
```

There seems to be a positive relationship between expenditure per student and average test scores. Also note that the points are dispersed, indicating that other factors may also influence test scores. We will further explore this using multiple regression in future sessions. 

For now, we will focus on a simple linear regression model with only one independent variable: expenditure per student.

Let's calculate the correlation coefficient between expenditure and test scores.

```{r}
with(cas, cor.test(expenditure, TestScore))
```


::: {.example #correlation}
What does the correlation coefficient tell us? Is it statistically significant? Justify your answer.
:::

<button class="solution-btn" onclick="myFunction('solution1')">Solution</button>
::: {#solution1 .solution-answer .env-green}

- **The reported correlation coefficient** is 0.19 between expenditure and TestScore.
  
  This is a positive but weak correlation: higher expenditures are associated with higher test scores, but the relationship is not very strong.

- **The p-value** = 7.989e-05, which is far below common significance levels (0.05, 0.01, or even 0.001).

  This means we reject the null hypothesis that the true correlation is zero.
  The result is statistically significant.
:::


## Simple Linear Regression

### Model Specification

We will fit a simple linear regression model to examine the relationship between expenditure per student and average test scores.

$$
\text{TestScore}_i = \beta_0 + \beta_1 \times \text{expenditure}_i + \varepsilon_i
$$ {#eq-linear-regression}


@eq-linear-regression is the linear regression model with a single independent variable.

Where:

:::{.callout-tip title="Key Components of the Linear Regression Model"}

- $\text{TestScore}_i$ in the Left Hand Side (LHS) is the **dependent variable** (response variable). $i$ indexes different school districts.
- $\text{expenditure}_i$ in the Right Hand Side (RHS) is the **independent variable** (explanatory variable or regressor).
- $\beta_0$ and $\beta_1$ are known as the parameters (coefficients) of the model.
  - $\beta_0$ is the **intercept**, representing the expected value of TestScore when expenditure is zero.
  - $\beta_1$ is the **slope coefficient**, representing the change in TestScore for a one dollar increase in expenditure.
- $\varepsilon_i$ is the error term, capturing all other factors affecting \text{expenditure} not included in the model, e.g, teacher quality, school facilities, parental involvement, etc.
:::


@eq-linear-regression can be written more generally as:

$$
Y_i = \beta_0 + \beta_1 X_i + \varepsilon_i
$$

- $\beta_1$ is the slope coefficient, which measures the change in the dependent variable $Y$ associated with a one-unit increase in the independent variable $X.$
- $\beta_0$ is the intercept, which is the expected value of $Y$ when $X=0;$ it is the point at which the population regression line intersects the Y axis. 
 
📌 In some econometric applications, the intercept has a meaningful economic interpretation. In other applications, the intercept has no real-world meaning; for example, when $X$ is the expenditure per student, strictly speaking the intercept is the expected value of test scores when there is no expenditure, which might be unrealistic. 

When the real-world meaning of the intercept is nonsensical, it is best to think of it simply as the coefficient that determines the level of the regression line.

### Estimating the Coefficients of the Linear Regression Model

Based on the scatter plot and the correlation analysis, we expect a positive relationship between expenditure and test scores. We will use the Ordinary Least Squares (OLS) method to estimate the coefficients of the linear regression model.

To run this regression, we use the `lm()` function in R.

```{r}
# Fit the linear regression model
model <- lm(TestScore ~ expenditure, data = cas)
summary(model)
```


### Interpretation of the output

#### A. **Model Specification and Fitted Values**

The regression estimates the effect of **expenditure** (per student, in USD) on **test scores** (combined reading and math).

**From Theory to Practice:**

We started with the theoretical model:
$$
\text{TestScore}_i = \beta_0 + \beta_1 \times \text{expenditure}_i + \varepsilon_i
$$

Using OLS estimation, we obtain the <span class="env-green">**fitted model**</span>:

$$
\widehat{\text{TestScore}} = 623.6 + 0.00575 \times \text{expenditure}
$$

The hat symbol ($\widehat{\text{TestScore}}$) indicates these are **predicted values** based on our sample data. Notice that we no longer have the error terms ($\varepsilon_i$) because this equation gives us the predicted values from the regression line.

**Understanding Residuals:**

Since our model can't perfectly predict every observation, we need to measure how far off our predictions are from reality:

$$
\text{Residual}_i = \text{TestScore}_i - \widehat{\text{TestScore}}_i
$$

Residuals represent the **difference between actual and predicted values** - essentially, they capture what our model couldn't explain. In the original theoretical model, these correspond to the error terms ($\varepsilon_i$) that we estimated.

---

#### B. **Coefficients**

* **Intercept (623.6):**
  When expenditure = 0, the predicted test score is about **623.6**.
  While not meaningful in practice (since expenditure cannot realistically be zero), it serves as the baseline of the regression line.

* **Expenditure (0.00575):**
  For every **one-dollar increase** in expenditure per student, the average test score is predicted to increase by **0.0057 points**.
  Put differently: a **[\$1,000]{.mathjax_ignore} increase** in expenditure is associated with a **5.75-point increase** in test scores.

---

#### C. **Statistical Significance**

For the slope coefficient ($\beta_1$):

* The **t-value = 3.984** and **p-value = 7.99e-05**, which is well below 0.001.
* This indicates that expenditure has a **statistically significant positive effect** on test scores.

---

#### D. **Model Fit**

* **R-squared = 0.0366** (about 3.7%).
  * This means expenditure explains only a **small portion of the variation** in test scores.
  * Many other factors (teacher quality, socio-economic background, school resources, etc.) likely play a larger role.

* **F-statistic = 15.87** with a p-value of 7.99e-05.
  * This tests the null hypothesis that all regression coefficients are equal to zero (no effect).
  * The low p-value indicates we reject the null hypothesis, confirming that the model has some explanatory power.


---

#### E. **Residuals**

The investigation of residuals requires diagnostic test which checks the assumptions of linear regression (e.g., homoscedasticity, normality of errors).

**We will cover them in future sessions**. 

Here we provide a brief overview.

```{r fig.asp=0.5, fig.width=8, echo=FALSE}
# Residual diagnostics
par(mfrow = c(1, 2))
plot(model, which = c(1,2))
```


* The first plot is **“Residuals vs Fitted” plot**. 
  
  The residuals appear reasonably balanced around zero, though the spread suggests substantial unexplained variation.

  The spread of residuals seems to **increase slightly with fitted values**, indicating potential heteroskedasticity (non-constant variance of errors). This suggests that the assumption of homoskedasticity may be violated, which could affect the reliability of our coefficient estimates and standard errors.

* The second plot is the **normal Q-Q plot**, which assesses whether the residuals are normally distributed. 

  * The points generally follow the reference line, suggesting that the residuals are approximately normally distributed, although there shows thin tails.
  
  * This indicates that there are few extreme residuals. The model predictions don't have large outliers.

**From the summary output:**

* The **residual standard error = 18.72**, which is the average size of prediction errors (in test score points).


---

✅ **Summary Interpretation:**

The regression analysis shows a **statistically significant positive relationship** between school expenditure and test scores. On average, an additional [\$1,000]{.mathjax_ignore} in per-student expenditure is associated with an increase of about **5.75 points** in test scores. However, the **explanatory power of the model is low (R² ≈ 3.7%)**, indicating that expenditure alone explains only a small fraction of the variation in test scores. This suggests that while spending matters, many other factors also influence student performance.

To gain a more comprehensive understanding of the factors affecting student achievement, future analyses should employ **multiple regression models** that incorporate additional explanatory variables such as student demographics, teacher qualifications, and school characteristics. We will cover multiple regression in subsequent sessions.


## Report by Stargazer

`stargazer` produces well-formatted regression tables that can be easily included in reports or publications.

```{r}
stargazer(model, type = "text", digits = 2)
```

--------------------------------------------------------------------------------

<h3>References</h3>

- [Introduction to Econometrics with R](https://www.econometrics-with-r.org/4.2-estimating-the-coefficients-of-the-linear-regression-model.html)