---
title: "Linear Model -- Lab2"
author: "Menghan Yuan"
date: "Sept 18, 2024"
output: 
    bookdown::html_document2:
        mathjax: "https://cdnjs.cloudflare.com/ajax/libs/mathjax/2.7.5/MathJax.js?config=TeX-AMS_CHTML.js"
        self_contained: false
        toc: true
        toc_float: true
        toc_depth: 4
        number_sections: false
        css: "style.css"
editor_options: 
  chunk_output_type: console
---

<SCRIPT language="JavaScript" SRC="my_jxscript.js"></SCRIPT>

<a class="top-link" href="#top" id="js-top">↑</a>

```{r setup, include=F, echo=F}
library(knitr) # load packages
library(kableExtra)
library(tidyverse)
library(latex2exp)
library(stargazer)
library(bookdown)
library(quantmod)

# don't show code unless we explicitly set echo = TRUE
opts_chunk$set(echo = TRUE, message=FALSE, fig.align="center", fig.pos = "H")
opts <- options(knitr.kable.NA = "")

## control long outputs by using eg `max.lines = 10`
hook_output_default <- knitr::knit_hooks$get('output')
truncate_to_lines <- function(x, n) {
   if (!is.null(n)) {
      x = unlist(stringr::str_split(x, '\n'))
      if (length(x) > n) {
         # truncate the output
         x = c(head(x, n), '...\n')
      }
      x = paste(x, collapse = '\n') # paste first n lines together
   }
   x
}
knitr::knit_hooks$set(output = function(x, options) {
   max.lines <- options$max.lines
   x <- truncate_to_lines(x, max.lines)
   hook_output_default(x, options)
})
```



```{r}
library(readxl)
cps_data <- read_excel("data/cps09mar.xlsx")
cps_data
```

Define two variables: experience and its square (we divide by 100 to simplify reporting).
Experience is defined as potential labor market experience:
$$
\begin{aligned}
\text{experience} &= \text{age}-\text{education}-6 \\
\text{exp2} &= \text{experience}^2/100
\end{aligned}
$$

```{r}
cps_data <- cps_data %>% 
    mutate(experience =  age-education-6,
           exp2=experience^2/100)
colnames(cps_data)
```

Calculate a new column for hourly wage using the following equation, and call it `wage`, then take the logarithm of it and call it `lwage`.
$$
\begin{aligned}
\text{wage} &= \frac{\text{earnings}}{\text{hours}\times\text{week}}  \\
\text{lwage} &= \log (\text{wage})
\end{aligned}
$$
```{r}
cps_data <- cps_data %>% 
    mutate(wage =  earnings/(hours*week),
           lwage = log(wage))
cps_data
```


___

# Simple Regression

For this illustration, we use the sub-sample of married (spouse present) black female wage earners with 12 years potential work experience. This sub-sample has 20 observations.
```{r}
sample1 <- cps_data %>% 
    filter(marital<3 & race==2 & female==1 & experience==12)
sample1
```

Now we want to estimate the following simple regression model:
$$
\log (\text{Wage}) = \beta_1 + \beta_2\; \text{education} + \varepsilon
$$

Manually calculated OLS using algebraic operations.

- matrix multiplication: `%*%`
- inverse of a matrix $A$: `solve(A)`
- inverse of matrix multiplication $A\times B$: `solve(A)%*%B` or `solve(A, B)`.

```{r}
y <- as.matrix(sample1$lwage)
x <- cbind(1, sample1[,"education"]) %>% as.matrix()
n <- length(y)
K <- ncol(x)
xx <- t(x) %*% x
xy <- t(x) %*% y
beta <- solve(xx,xy)
beta
```

Variance matrix, $V_{\hat{\beta}}$
```{r}
e <- y-x%*%beta
sigma_hat <- sum(e^2)/(n-K)
v0 <- sigma_hat * solve(xx)
s0 <- diag(v0) %>% sqrt()
s0
```

We can write the estimated equation as:
$$
\widehat{\log \text{(Wage)}} = 0.698 + 0.155\; \text{education} .
$$

Compare your estimate with the output from `lm` function.
```{r}
lm_model1 <- lm(lwage~education, data=sample1)
summary(lm_model1)
```


___

# Multivariate Regression

Now let's carry out a multivariate regression model by introducing two additional regressors: experience and its square ($\text{experience}^2/100$, we divide by 100 to simplify reporting).

This time, we use the sub-sample of single (never married) Asian men, which has 268 observations.
```{r}
sample2 <- cps_data %>% filter(race==4 & marital==7 & female==0)
sample2
```


```{r}
lm_model2 <- lm(lwage~education+experience+exp2, data=sample2)
summary(lm_model2)
```



We obtain the estimates:
$$
\widehat{\log \text{(Wage)}} = 0.575 + 0.143\; \text{education} + 0.036 \; \text{experience} - 0.071 \; \text{experience}^2/100.
$$

$$
\sum_{i=1}^n e_i
$$

```{r}
e <- lm_model2$residuals
sum(e)
```


Let $x_{2i}$ be education and $x_{3i}$ be experience, calculate 
$$
\begin{aligned}
&\sum_{i=1}^n x_{2i}\, e_i \\
&\sum_{i=1}^n x_{3i}\, e_i \quad \text{, and} \quad \\
&\sum_{i=1}^n x_{3i}^2\, e_i
\end{aligned}
$$
```{r}
sum(sample2$education * e)
sum(sample2$experience * e)
sum(sample2$exp2 * e)
```

Calculate 
$$
\sum_{i=1}^n \hat{y}_i \,e_i
$$

```{r}
sum(lm_model2$fitted.values * e)
```


Visualize residuals.

1. "Residuals vs Fitted" plot 
```{r}
plot(lm_model2, which=1)
```

2. "Residual Q-Q" plot 
```{r}
plot(lm_model2, which=2)
```









