---
title: "Regression with a Binary Dependent Variable"
---

::: {.callout-note appearance="simple" icon=false}
🎯 **Study Objectives**

- Understand why linear regression is inappropriate for binary dependent variables.
- Learn the logit and probit models for analyzing binary outcomes.
- Interpret coefficients in logistic regressions.
- Apply logit/probit models using real-world data (HMDA mortgage application dataset).
- Understand the concept of predicted probabilities.
- Compare linear probability model (LPM), logit, and probit approaches.
:::

## Why Binary Dependent Variables Require Special Treatment

In many economic and financial applications, the dependent variable is **binary** (takes only two values: 0 or 1). Examples include:

- **Mortgage approval**: Did the bank approve the mortgage application? (Yes = 1, No = 0)
- **Labor force participation**: Is the individual employed? (Yes = 1, No = 0)
- **Default risk**: Did the borrower default on the loan? (Yes = 1, No = 0)
- **Market entry**: Does the firm enter the market? (Yes = 1, No = 0)
- **Product choice**: Does the consumer purchase the product? (Yes = 1, No = 0)

### Why Linear Regression Fails

If we try to use ordinary linear regression (OLS) with a binary dependent variable $Y_i \in \{0,1\}$:

$$
Y_i = \beta_0 + \beta_1 X_{1i} + \beta_2 X_{2i} + \cdots + \beta_k X_{ki} + u_i
$$

This approach is called the **Linear Probability Model (LPM)** because it estimates the probability that $Y_i = 1$ as a linear function of the predictors.

**Problems with LPM:**

1. **Predicted probabilities outside [0,1]**: OLS can produce fitted values $\hat{Y}_i < 0$ or $\hat{Y}_i > 1$, which makes no sense for probabilities.

2. **Heteroskedasticity**: The error term variance depends on $X$, violating the homoskedasticity assumption. This means standard errors are incorrect.

3. **Nonlinear relationship**: The effect of $X$ on the probability of $Y=1$ is unlikely to be constant across all values of $X$. 
   For example, increasing income from [\$20,000]{.mathjax_ignore} to [\$30,000]{.mathjax_ignore} likely has a different effect on mortgage approval than increasing from [\$200,000]{.mathjax_ignore} to [\$210,000]{.mathjax_ignore}.

### The Solution: Logit and Probit Models

To address these problems, we use **nonlinear models** that ensure predicted probabilities stay within [0,1]:

- **Logit model**: Uses the logistic cumulative distribution function (CDF)
- **Probit model**: Uses the standard normal cumulative distribution function (CDF)


## The HMDA Dataset: Mortgage Lending Discrimination

We will use data from the **Federal Reserve Bank of Boston** collected under the **Home Mortgage Disclosure Act (HMDA)**. The dataset contains information on mortgage applications in the Boston area in 1990. The dataset contains 2380 observations with 14 variables.

**Research Question:** Do banks discriminate against minority applicants when making mortgage lending decisions, after controlling for economic factors?

### Load the Data

```{r}
# Load required packages
pkgs <- c("AER", "tidyverse", "stargazer", "margins", "data.table")
missing <- setdiff(pkgs, rownames(installed.packages()))
if (length(missing) > 0) install.packages(missing)
invisible(lapply(pkgs, function(pkg) suppressPackageStartupMessages(library(pkg, character.only = TRUE))))

# Load HMDA data
data("HMDA")

# Preview the data
data.table(HMDA)
```

### Variable Definitions

The key variables in the HMDA dataset are:

- We will mainly use `deny`, `pirat`, and `afam` in our analysis.

| Variable | Definition |
|----------|------------|
| <span class="env-green">`deny`</span> | Was the mortgage application denied? (yes/no) — **dependent variable** |
| <span class="env-green">`pirat`</span> | Payments-to-income ratio (monthly loan payments / monthly income) |
| <span class="env-green">`afam`</span> | Is the applicant African American? (yes/no) |
| `hirat` | Housing expense-to-income ratio (monthly housing expenses / monthly income) |
| `lvrat` | Loan-to-value ratio (loan amount / assessed property value) |
| `chist` | Credit history: consumer credit score |
| `mhist` | Mortgage history: mortgage credit score |
| `phist` | Public record history: coded as "no" or "yes" (any bankruptcies, tax liens, etc.) |
| `insurance` | Was mortgage insurance denied? (yes/no) |
| `selfemp` | Is the applicant self-employed? (yes/no) |
| `single` | Is the applicant single? (yes/no) |
| `hschool` | Does the applicant have high school education? (yes/no) |
| `unemp` | 1989 Massachusetts unemployment rate in applicant's industry (%). |
| `condominium` | Is the property a condominium? (yes/no) |


### Data Exploration

#### Frequency Tables and Contingency Tables

**Let's examine the denial rates overall and by race.**

We first look at the overall denial counts and rates.

```{r results='hold'}
# Overall denial counts
cat("Overall Denial Counts:\n")
with(HMDA, table(deny))

# Overall denial rate
cat("===========================\n")
cat("Overall Denial Rate:\n")
with(HMDA, prop.table(table(deny))) %>% round(3)
```


The table shows that about 12% of mortgage applications were denied overall. The majority of applications (88%) were approved.

Next, we examine denial counts and rates by race.


A <span class="env-green">**contingency table**</span> is an effective method to see the association between two categorical variables. It counts the number of observations in each of the four possible scenarios.
When dealing with just one categorical variable, this is referred to as a **frequency table**, which count the number of observations for each category.

The following gives a 2x2 contingency table for mortgage denial by African-American or not.

```{r results='hold'}
# Denial rate by race
cat("Denail counts by race:\n")
with(HMDA, table(deny, afam))
cat("===========================\n")
cat("Denial rate by race:\n")
with(HMDA, prop.table(table(deny, afam), margin = 2)) %>% round(3)
```


Some observations from the table:

- The majority of applicants are non-African American.
- African American applicants have a higher denial rate (about 28%) compared to non-African American applicants (about 9%).

:::{.callout-note appearance="simple" icon=false}
This descriptive evidence suggests that *the likelihood of denial may be systematically higher for African American applicants*. However, these simple proportions do not control for other relevant factors such as income, credit history, or loan-to-value ratio. 
Logistic regression will allow us to model this relationship more rigorously while accounting for these additional variables.
:::

#### Summary Statistics

```{r}
summary(HMDA)
```

Summary statistics by race.

```{r}
# Summary statistics by race
HMDA %>%
    group_by(afam) %>%
    summarise(
        n = n(),
        denial_rate = mean(deny == "yes"),
        mean_pirat = mean(pirat),
        mean_lvrat = mean(lvrat)
    )
```

### Visualizing the Contingency Table

We can visualize the contingency table using a **Stacked Bar Plot**.

```{r echo=FALSE}
# Create contingency table
contingency_data <- HMDA %>%
    select(deny, afam) %>%
    mutate(
        afam = factor(afam, levels = c("no", "yes"), labels = c("Non-African American", "African American")),
        deny = factor(deny, levels = c("no", "yes"), labels = c("Approved", "Denied"))
    )

contingency_table <- with(contingency_data, table(deny, afam))

# Stacked bar plot
contingency_table %>%
    as_tibble() %>%
    mutate(
        # fix order of factors
        deny = factor(deny, levels = c("Approved", "Denied")),
        afam = factor(afam, levels = c("Non-African American", "African American"))
        ) %>%
    ggplot(aes(x = afam, y = n, fill = deny)) +
    geom_bar(position = "stack", stat = "identity", color = "black", linewidth = 0.3) +
    scale_fill_manual(values = c("Approved" = "#88CCEE", "Denied" = "#CC6677")) +
    labs(
        x = "",
        y = "Frequency",
        fill = "Application Status",
        title = "Mortgage Application Outcomes by Race"
    ) +
    theme_minimal(base_size = 14) +
    theme(
        legend.position = "bottom",
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 12),
        axis.text = element_text(size = 12),
        plot.title = element_text(hjust = 0.5)
    )
```

The stacked bar graph shows:

- The sample sizes of African American and non-African American applicants
- The distribution of approved vs. denied applications within each racial group
- African American applicants appear to have a higher proportion of denials


Issue: When the groups have very different sizes, it can be hard to compare proportions using absolute frequencies.

Remedy: Use a **mosaic plot** to visualize **relative frequencies**.

#### Mosaic Plot

A mosaic plot replaces absolute frequencies with relative frequencies, making it easier to compare proportions across groups.

```{r echo=FALSE}
# Mosaic plot
mosaicplot(contingency_table %>% t(),
    xlab = "Race", ylab = "Application Status",
    main = "Mosaic Plot: Race × Mortgage Denial",
    color = c("#88CCEE", "#CC6677"),
    cex.axis = 1
)
```

**Interpretation:**

- The widths of the boxes are proportional to the percentage of each racial group in the sample.
- The heights represent the denial rates within each group.
- We can see that the "Denied" box for African American applicants is taller than for non-African American applicants, indicating a higher denial rate.

### Measures of Risk and Association for Binary Outcomes

To quantify the difference in mortgage denial rates between racial groups, we can calculate several measures of risk and association.

```{r results='hold'}
# Calculate risk (denial rate) for each group
risk_table <- contingency_table %>%
    prop.table(margin = 2) %>%
    as.data.frame.matrix()
risk_table %>% round(3)
# Extract probabilities
p_non_afam <- risk_table["Denied", "Non-African American"] # P(deny=1 | afam=0)
p_afam <- risk_table["Denied", "African American"] # P(deny=1 | afam=1)

cat("\nRisk (Denial Rate) by Race:\n")
cat("===========================\n")
cat(sprintf("Non-African American: %.4f (%.2f%%)\n", p_non_afam, p_non_afam * 100))
cat(sprintf("African American:     %.4f (%.2f%%)\n", p_afam, p_afam * 100))
```

1. **Risk Difference (Excess Risk)**

   The **risk difference** or **excess risk** (ER) is the difference in denial rates between the two groups:

   $$
   ER = P(\text{deny}=1|\text{afam}=1) - P(\text{deny}=1|\text{afam}=0)
   $$

   ```{r results='hold'}
   ER <- p_afam - p_non_afam
   cat("\nRisk Difference (Excess Risk):\n")
   cat("==============================\n")
   cat(sprintf("ER = %.4f - %.4f = %.4f (%.2f percentage points)\n", p_afam, p_non_afam, ER, ER * 100))
   ```

   **Interpretation:** African American applicants have a denial rate that is `r round(ER*100, 2)` percentage points higher than non-African American applicants.

   - If $ER = 0$: no difference in risk between groups
   - If $ER > 0$: higher risk for the <span class="env-green">treatment group</span> (African Americans, `afam=1`)
     
     By contrast, the <span class="env-green">control group</span> is non-African Americans (`afam=0`).
   - If $ER < 0$: lower risk for the treatment group

2. **Risk Ratio (Relative Risk)**

   The **risk ratio** or **relative risk** (RR) is the ratio of denial rates:

   $$
   RR = \frac{P(\text{deny}=1|\text{afam}=1)}{P(\text{deny}=1|\text{afam}=0)}
   $$

   ```{r results='hold'}
   RR <- p_afam / p_non_afam
   cat("\nRisk Ratio (Relative Risk):\n")
   cat("===========================\n")
   cat(sprintf("RR = %.4f / %.4f = %.4f\n", p_afam, p_non_afam, RR))
   ```

   **Interpretation:** African American applicants have a denial rate that is `r round(RR, 2)` times higher than non-African American applicants.

   - If $RR = 1$: no difference in risk
   - If $RR > 1$: higher risk for the treatment group (African Americans, `afam=1`)
   - If $RR < 1$: lower risk for the treatment group

3. **Odds Ratio**

   The **odds ratio** (OR) compares the odds of denial between the two groups.

   First, calculate the odds for each group:

   $$
   \text{odds} = \frac{P(\text{deny}=1)}{P(\text{deny}=0)} = \frac{P(\text{deny}=1)}{1 - P(\text{deny}=1)}
   $$

   ```{r results='hold'}
   # Calculate odds for each group
   odds_non_afam <- p_non_afam / (1 - p_non_afam)
   odds_afam <- p_afam / (1 - p_afam)

   cat("\nOdds by Race:\n")
   cat("=============\n")
   cat(sprintf("Non-African American: %.4f\n", odds_non_afam))
   cat(sprintf("African American:     %.4f\n", odds_afam))
   ```
   
   This means that the odds of being denied a mortgage are approximately **0.10 for non--African American applicants** and **0.40 for African American applicants**.
   
   The <span class="env-green">**odds ratio**</span> is:

   $$
   OR = \frac{\text{odds}(\text{afam}=1)}{\text{odds}(\text{afam}=0)} = \frac{P(\text{deny}=1|\text{afam}=1)/[1-P(\text{deny}=1|\text{afam}=1)]}{P(\text{deny}=1|\text{afam}=0)/[1-P(\text{deny}=1|\text{afam}=0)]}
   $$

   ```{r results='hold'}
   OR <- odds_afam / odds_non_afam
   cat("\nOdds Ratio:\n")
   cat("===========\n")
   cat(sprintf("OR = %.4f/%.4f = %.4f\n", odds_afam, odds_non_afam, OR))
   ```

   **Interpretation:** The odds of mortgage denial for African American applicants are `r round(OR, 2)` times higher than the odds for non-African American applicants.

   - If $OR = 1$: no difference in odds
   - If $OR > 1$: higher odds for the treatment group (African Americans, `afam=1`)
   - If $OR < 1$: lower odds for the treatment group

:::{.callout-note}
## Important Distinction

- **Risk difference** measures the absolute difference in probabilities (additive scale)
- **Risk ratio** and **odds ratio** measure relative differences (multiplicative scale)
- The odds ratio is particularly useful because <span class="env-green">it directly relates to the coefficient in logistic regression</span>: 
  
  When we estimate a logistic model with `afam` as the predictor, $e^{\beta_{\text{afam}}}$ will approximately equal the odds ratio we calculated here.
:::

## Model Specification: The Boston Mortgage Example

We want to model the probability that a mortgage application is denied as a function of applicant characteristics.

### Preparing the Data

First, we create a binary numeric variable for denial (1 = denied, 0 = approved) and recode some factors:

```{r results='hold'}
# Create binary dependent variable
HMDA$deny_binary <- ifelse(HMDA$deny == "yes", 1, 0)

# Create binary variable for African American
HMDA$afam_binary <- ifelse(HMDA$afam == "yes", 1, 0)

# Check the variables
head(HMDA[, c("deny", "deny_binary", "pirat", "afam", "afam_binary")])
```

### Model 1: Linear Probability Model

Let's start with a simple linear regression to see its limitations.

The OLS regression of the binary dependent variable, $deny$ against the payment-to-income ratio (`pirat`), is estimated as:

$$
deny = \beta_0 + \beta_1 \cdot pirat + \varepsilon
$$

```{r}
lpm_simple <- lm(deny_binary ~ pirat, data = HMDA)
coeftest(lpm_simple, vcov. = vcovHC, type = "HC1")
```

$$
\widehat{deny} = -0.080 + 0.604 \cdot pirat
$$

The estimated coefficient on P/I ratio is positive, and the population coefficient is statistically significantly different from 0 at the 1% level (the t-statistic is 6.12). 
If P/I ratio increases by 0.1, the probability of denial increases by approximately $0.604\times 0.1 \approx 0.060,$ that is, by 6 percentage points.
Thus applicants with higher debt payments as a fraction of income are more likely to have their application denied.

Now we plot the data and the regression line to visualize the model.

```{r #fig-linear, echo=FALSE}
# plot the data
plot(
    x = HMDA$pirat,
    y = HMDA$deny_binary,
    main = "Linear Model of Mortgage Application Denial, Given the Payment-to-Income Ratio",
    xlab = "P/I ratio",
    ylab = "Deny",
    pch = 20,
    ylim = c(-0.4, 1.4),
    cex.main = 0.8
)

# add horizontal dashed lines and text
abline(h = 1, lty = 2, col = "darkred")
abline(h = 0, lty = 2, col = "darkred")
text(2.5, 0.9, cex = 0.8, "Mortgage denied")
text(2.5, -0.1, cex = 0.8, "Mortgage approved")

# add the estimated regression line
abline(lpm_simple,
    lwd = 1.8,
    col = "steelblue"
)
```

::: {.callout-warning title="Shortcomings with the linear probability model" icon="false"}

1. The linear probability model can predict probabilities outside the $[0,1]$ range. For example, for high values of P/I ratio (`pirat`), the predicted probability exceeds 1. However, probabilities must lie between 0 and 1.
2. The relationship between P/I ratio and the probability of denial may NOT be linear in reality. 
   
   It is reasonable to expect the marginal effects of P/I ratio on denial probability to diminish as P/I ratio increases. 
   
   Although a change in P/I ratio from 0.3 to 0.4 might have a large effect on the probability of denial, once the P/I ratio is already very high (e.g., 0.9 to 1.0), increasing P/I ratio further will have litte effect.
3. The error term in the linear probability model is *heteroskedastic*, violating OLS assumptions. 
   
   This means that standard errors and hypothesis tests based on OLS are invalid. → This issue can be addressed using heteroskedasticity robust standard errors.
:::

### Model 2: Logit Regression with a single predictor

The estimated model is 
$$
\mathrm P(deny_i = 1 | pirat_i) = F(\beta_0 + \beta_1 \cdot pirat_i) = \frac{1}{1 + e^{-(\beta_0 + \beta_1 \cdot pirat_i)}} ,
$$

where $F$ is the logistic distribution function.

Note that the left-hand side is the probability that the $i$-th mortgage application is denied ($deny_i = 1$), given the payment-to-income ratio ($pirat_i$).

The strength of the logit model is that <span class="env-green">it ensures predicted probabilities are always between 0 and 1</span>.

```{r}
# Estimate logit model
logit_simple <- glm(deny_binary ~ pirat,
    family = binomial(link = "logit"),
    data = HMDA
)
coeftest(logit_simple, vcov. = vcovHC, type = "HC1")
```

$$
\widehat{\mathrm P}(deny_i = 1 | pirat_i) = F(-4.02 + 5.88 \cdot pirat_i)= \frac{1}{1 + e^{-(-4.02 + 5.88 \cdot pirat_i)}}
$$

Visualizing the logit model:

```{r #fig-logit, echo=FALSE, fig.cap="The logit model uses the logistic distribution function to model the probability of denial as a function of the payment-to-income ratio. Unlike the linear probability model, the logit model ensures predicted probabilities remain within the [0,1] range."}
# plot data
plot(
    x = HMDA$pirat,
    y = HMDA$deny_binary,
    main = "Logit Model of the Probability of Denial, Given the Payment-to-Income Ratio",
    xlab = "P/I ratio",
    ylab = "Deny",
    pch = 20,
    ylim = c(-0.4, 1.4),
    cex.main = 0.85
)

# add horizontal dashed lines and text
abline(h = 1, lty = 2, col = "darkred")
abline(h = 0, lty = 2, col = "darkred")
text(2.5, 0.9, cex = 0.8, "Mortgage denied")
text(2.5, -0.1, cex = 0.8, "Mortgage approved")

# add estimated regression line
x <- seq(0, 3, 0.01)
y <- predict(logit_simple, list(pirat = x), type = "response")

lines(x, y, lwd = 1.5, col = "steelblue")
```

The estimated logit regression function has a stretched “S” shape: It is nearly 0 and flat for small values of P/I ratio, it turns and increases for intermediate values, and it flattens out again and is nearly 1 for large values. 

::: {.example #exm-1}
**Predicted probability.**

What is the probability of denial given a P/I ratio of 0.3? What about for P/I ratio being 0.4 and 0.5?
:::

<button class="solution-btn" onclick="myFunction('sol-1')">Solution</button>
<div id="sol-1" class="solution-answer env-green">
For P/I ratio of 0.3, the estimated probability of denial based on the estimated logit model is:
$$
\widehat{P}(deny = 1 | pirat = 0.3) = \frac{1}{1 + e^{-(-4.02 + 5.88 \cdot 0.3)}} \approx 9.4\%
$$
That is, the probability of denial is approximately 9.4%.

For P/I ratio of 0.4:
$$
\widehat{P}(deny = 1 | pirat = 0.4) = \frac{1}{1 + e^{-(-4.02 + 5.88 \cdot 0.4)}} \approx 15.7\%
$$
That is, the probability of denial is approximately 15.7%.

For P/I ratio of 0.5:
$$
\widehat{P}(deny = 1 | pirat = 0.5) = \frac{1}{1 + e^{-(-4.02 + 5.88 \cdot 0.5)}} \approx 25.2\%
$$

**Main takeaway:** As P/I ratio increases, the probability of denial increases, but not linearly.
</div>


<span class="env-green">**Interpretation on the logit coefficients:**</span>

The estimated model is

$$
\widehat{P}(deny_i = 1 \mid pirat_i) = \frac{1}{1 + e^{-(-4.02 + 5.88 \cdot pirat_i)}}
$$

which can be written in **log-odds form** as

$$
\log\left(\frac{P(deny_i = 1)}{1 - P(deny_i = 1)}\right) = -4.02 + 5.88 \cdot pirat_i
$$

1. **Interpreting coefficients (log-odds scale):**
   
   The slope coefficient, $\hat{\beta}_1 = 5.88$, means that a **one-unit increase** in `pirat` increases the **log-odds** of mortgage denial by **5.88**, holding all else constant. 

2. **Interpreting odds ratios (exponentiating coefficients):**

   To obtain a more intuitive interpretation, we exponentiate the coefficient:
   $$
   e^{\hat{\beta}_1} = e^{5.88} \approx 357.7
   $$
   This means that for a **one-unit increase** in the payment-to-income ratio, the **odds** of mortgage denial are about **357.7 times larger**.
   
   - Because a one-unit increase in `pirat` is very large in practice, it is often more meaningful to interpret smaller changes.
 
   For example, for a **0.1 increase** in `pirat`:
   $$
   e^{5.88 \times 0.1} \approx 1.80
   $$
   This means that a 0.1 increase in the payment-to-income ratio increases the **odds of denial by about 80%**.

**Hypothesis Testing:**

Using the normal distribution of parameter estimates, we can use the standard normal table rather than the $t$ table for critical points to test hypotheses about the coefficients.

The $z$-statistic for testing $H_0: \beta_1 = 0$ is:

$$
z = \frac{\hat{\beta}_1 - 0}{SE(\hat{\beta}_1)} = \frac{5.88 - 0}{1} \approx 5.88
$$

Since $z > 1.96$, we reject the null hypothesis at the 5% significance level and conclude that there is a statistically significant positive relationship between payment-to-income ratio and the probability of mortgage denial.

### Model 3: Linear Probability Model with Multiple Predictors
$$
deny_i = \beta_0 + \beta_1 \cdot pirat_i + \beta_2 \cdot afam_i
$$

```{r}
# Estimate LPM
lpm_multi <- lm(deny_binary ~ pirat + afam_binary, data = HMDA)
summary(lpm_multi)
```

**Interpretation:**

- $\hat{\beta}_1$: A one-unit increase in the payment-to-income ratio increases the probability of denial by approximately $\hat{\beta}_1$.
- $\hat{\beta}_2$: African American applicants have a probability of denial that is $\hat{\beta}_2$ percentage points higher than non-African American applicants, holding `pirat` constant.

### Model 4: Logit Model with Multiple Predictors

Now we estimate a logit model including both P/I ratio (`pirat`) and African-American binary (`afam_binary`) as predictors:

$$
P(deny_i = 1 | pirat_i, afam_i) = F(\beta_0 + \beta_1 \cdot pirat_i + \beta_2 \cdot afam_i) = \frac{1}{1 + e^{-(\beta_0 + \beta_1 \cdot pirat_i + \beta_2 \cdot afam_i)}} .
$$


```{r}
# Estimate logit model
logit_multi <- glm(deny_binary ~ pirat + afam_binary,
    family = binomial(link = "logit"),
    data = HMDA
)
coeftest(logit_multi, vcov. = vcovHC, type = "HC1")
```


The estimated logit model is:
$$
\begin{split}
\widehat{P}(deny_i = 1 | pirat_i, afam_i) &= F(-4.13 + 5.37 \cdot pirat_i + 1.27 \cdot afam_i) \\
&= \frac{1}{1 + e^{-(-4.13 + 5.37 \cdot pirat_i + 1.27 \cdot afam_i)}}
\end{split}
$$

<span class="env-green">**Interpretation of Logit Coefficients**</span>

The coefficients in a logit model represent the change in the **log-odds** of denial:

$$
\log\left(\frac{P(deny = 1)}{1 - P(deny = 1)}\right) = -4.13 + 5.37 \cdot pirat_i + 1.27 \cdot afam_i
$$

**1. Interpreting coefficients (log-odds scale):**

- $\hat{\beta}_1 > 0$: A one-unit increase in `pirat` increases the log-odds of denial by $\hat{\beta}_1$, holding other variables constant.
- $\hat{\beta}_2 > 0$: Being African American increases the log-odds of denial by $\hat{\beta}_2$ compared to non-African Americans, holding other variables constant.

**2. Interpreting odds ratios (exponentiating coefficients):**

Alternatively, we can interpret the **odds ratio** by taking the exponential of the coefficients:

```{r include=FALSE}
# Calculate odds ratios
exp(coef(logit_multi)) %>% round(3)
```

- **For `pirat`**: $e^{\hat{\beta}_1}$ represents the **multiplicative change in odds** for a one-unit increase in the payment-to-income ratio. 
  
  For example, if the P/I ratio increases by 0.2, then $e^{5.37\times 0.2} = 2.93$, that is, the odds of mortgage denial are approximately 2.94 times higher, holding `afam` constant.

- **For `afam`**: $e^{\hat{\beta}_2}$ is the **odds ratio comparing African American applicants to non-African American applicants**.
  
  Based on the estimates, $e^{1.2} = 3.56$, that is, African American applicants have odds of denial that are 3.56 times higher than non-African American applicants, holding `pirat` constant.

::: {.example #exm-2}

Suppose we have:

- Odds of denial for a non-African American applicant with `pirat = 0.3` is 0.15
- The odds ratio for African Americans compared to non-African American is $e^{\hat{\beta}_2} = 3.56$

Calculate the **expected odds** for an African American applicant with the same `pirat = 0.3`.
:::

<button class="solution-btn" onclick="myFunction('sol-2')">Solution</button>
::: {#sol-2 .solution-answer .env-green}
$$
\text{odds}(\text{afam}=1) = \text{odds}(\text{afam}=0) \times e^{\hat{\beta}_2} = 0.15 \times 3.56 = 0.534
$$
The **expected odds** of denial for an African American applicant with `pirat = 0.3` is 0.534.
:::


### Model 5: Probit Model

In contrast to the logit model, which uses the logistic CDF, the probit model uses the <span class="env-green">standard normal</span> distribution function to model the probability of denial.
$$
P(deny_i = 1 | X_i) = \Phi(\beta_0 + \beta_1 \cdot pirat_i + \beta_2 \cdot afam_i)
$$

where $\Phi(\cdot)$ is the standard normal CDF.

```{r}
# Estimate probit model
probit_multi <- glm(deny_binary ~ pirat + afam_binary,
    family = binomial(link = "probit"),
    data = HMDA
)
coeftest(probit_multi, vcov. = vcovHC, type = "HC1")
```

### Comparison: Logit, and Probit
```{r echo=FALSE}
# plot data
plot(
    x = HMDA$pirat,
    y = HMDA$deny_binary,
    main = "Probit and Logit Models Model of the Probability of Denial, Given P/I Ratio",
    xlab = "P/I ratio",
    ylab = "Deny",
    pch = 20,
    ylim = c(-0.4, 1.4),
    cex.main = 0.9
)

# add horizontal dashed lines and text
abline(h = 1, lty = 2, col = "darkred")
abline(h = 0, lty = 2, col = "darkred")
text(2.5, 0.9, cex = 0.8, "Mortgage denied")
text(2.5, -0.1, cex = 0.8, "Mortgage approved")

# add estimated regression line of Probit and Logit models
x <- seq(0, 3, 0.01)
y_probit <- predict(probit_multi, list(pirat = x, afam_binary = rep(0, 301)), type = "response")
y_logit <- predict(logit_multi, list(pirat = x, afam_binary = rep(0, 301)), type = "response")

lines(x, y_probit, lwd = 1.5, col = "steelblue")
lines(x, y_logit, lwd = 1.5, col = "black", lty = 2)

# add a legend
legend("topleft",
    horiz = TRUE,
    legend = c("Probit", "Logit"),
    col = c("steelblue", "black"),
    lty = c(1, 2)
)
```

**Practical Recommendation:**

- Logit and probit typically give very similar results.
- Logit is more common in economics and finance due to the convenient odds ratio interpretation.

### Predicted Probabilities

To compare effects across models, we calculate **predicted probabilities** at specific values of the predictors.

::: {.example #exm-3}
For the three multivariate models (models 3–5), what is the predicted probability of denial for:

- An African American applicant with `pirat = 0.3`?
- A non-African American applicant with `pirat = 0.3`?
:::

<button class="solution-btn" onclick="myFunction('sol-3')">Solution</button>
::: {#sol-3 .solution-answer .env-green}
&nbsp;

**For linear probability model (LPM):**

- African American applicant (`afam_binary = 1`):
  $$
  \widehat{deny} = -0.09 + 0.56 \times 0.3 + 0.18 \times 1 = 0.258 
  $$
- Non-African American applicant (`afam_binary = 0`):
  $$
  \widehat{deny} = -0.09 + 0.56 \times 0.3 + 0.18 \times 0 = 0.078
  $$

**For logit model:**

- African American applicant:
  $$
  \widehat{P}(deny = 1 | pirat = 0.3, afam = 1) = \frac{1}{1 + e^{-(-4.13 + 5.37 \times 0.3 + 1.27 \times 1)}} \approx 0.22
  $$
- Non-African American applicant:
  $$
  \widehat{P}(deny = 1 | pirat = 0.3, afam = 1) = \frac{1}{1 + e^{-(-4.13 + 5.37 \times 0.3 + 1.27 \times 0)}} \approx 0.07
  $$

**For probit model:**

- African American applicant:
  $$
  \widehat{P}(deny = 1 | pirat = 0.3, afam = 1) = \Phi(-2.26 + 2.74 \times 0.3 + 0.71 \times 1) = \Phi(-0.728)
  $$
  Refer to the standard normal table for $\Phi(-0.728) \approx 0.23$.
- Non-African American applicant:
  $$
  \widehat{P}(deny = 1 | pirat = 0.3, afam = 0) = \Phi(-2.26 + 2.74 \times 0.3 + 0.71 \times 0) \approx 0.075
  $$
  Refer to the standard normal table for $\Phi(-1.438) \approx 0.075$.
:::

```{r validate, include=FALSE}
# Create prediction data
newdata <- data.frame(
    pirat = c(0.3, 0.3),
    afam_binary = c(1, 0)
)

# Predictions from each model
pred_lpm <- predict(lpm_multi, newdata = newdata, type = "response")
pred_logit <- predict(logit_multi, newdata = newdata, type = "response")
pred_probit <- predict(probit_multi, newdata = newdata, type = "response")

# Compare predictions
data.frame(
    Applicant = c("African American", "Non-African American"),
    LPM = pred_lpm,
    Logit = pred_logit,
    Probit = pred_probit
)
```



## Summary: Logit vs. Probit vs. LPM

|  | Linear | Logit | Probit |
|---------|-----|-------|--------|
| **Function** | Linear | Logistic CDF | Normal CDF |
| **Predicted probabilities** | Can be outside [0,1] | Always in [0,1] | Always in [0,1] |
| **Interpretation** | Direct (probability) | Log-odds / Odds ratio | Z-score |
| **Heteroskedasticity** | Always present | Accounted for | Accounted for |
| **When to use** | Quick approximation | Preferred for most applications | Similar to logit; standard in some fields |



